Parsing Postfix log files

### 2009-03-09 doc/logparser-acronyms.tex
Corrections based on comments from Carl, plus a few I noticed in the process.

\newglossaryentry{CFG}{
    type=\acronymtype,
    name={CFG},
    description={Context Free Grammar},
    first={Context Free Grammar (CFG)}
}
\newglossaryentry{CFG glossary}{
    name={Context Free Grammar},
    description={
        See Context Free Language.
    }
}

\newglossaryentry{CFL}{
    type=\acronymtype,
    name={CFL},
    description={Context Free Language},
    first={Context Free Language (CFL)}
}
\newglossaryentry{CFL glossary}{
    name={Context Free Language},
    description={
        A Context Free Language is a language generated or parsed by a
        \acronym{CFG}.  Every transition in a CFG takes the form
        $L\rightarrow{}R$, where $L$ is a single non-terminal symbol, and
        $R$ is a (possibly empty) string of terminal and/or non-terminal
        symbols.  Every CFG can be recognised by a \acronym{PDA}, and
        \acronyms{PDA} cannot recognise more complicated grammars, so
        \acronyms{CFL}, \acronyms{CFG}, and \acronyms{PDA} are equivalent.
    }
}

>>>>

\newglossaryentry{CFG}{
    type=\acronymtype,
    name={CFG},
    description={Context Free Grammar},
    first={Context Free Grammar (CFG)}
}
\newglossaryentry{CFG glossary}{
    name={Context Free Grammar},
    description={
        A CFG is a grammar in which every transition or production takes
        the form $L\rightarrow{}R$, where $L$ is a single non-terminal
        symbol, and $R$ is a (possibly empty) string of terminal and/or
        non-terminal symbols.  Every CFG can be recognised by a
        \acronym{PDA}, and \acronyms{PDA} cannot recognise more complicated
        grammars, so \acronyms{CFG} and \acronyms{PDA} are equivalent.
    }
}

### 2009-03-09 doc/logparser-background.tex
Corrections based on comments from Carl, plus a few I noticed in the process.

\begin{table}[ht]
    \caption{Sample output from SQL query}
    \empty{}\label{Sample output from SQL query}
    \begin{tabular}[]{lrr}
        \tabletopline{}%
        Restriction & Number of  & Percentage of    \\
                    & rejections & total rejections \\
        \tablemiddleline{}%
        \input{build/include-sample-sql-output}
        \tablebottomline{}%
    \end{tabular}
\end{table}

@@

This example client is so poorly written that not only does it present the
sender and recipient addresses improperly, it ignores the error messages
returned by the server and carries on regardless.  Many spam and virus
sending programs have serious deficiencies; unfortunately other programs,
particularly newer programs, were written by competent programmers or send
mail using well written programs (e.g.\ Postfix or Sendmail on Unix hosts,
Microsoft Outlook on Windows hosts).  Traditionally a mail server would
have done its best to deal with deficient clients, with the intention of
accepting as much mail sent to its users as possible, e.g.\ by ignoring the
absence of a HELO command, or by accepting sender or recipient addresses
that were not enclosed in \texttt{<>}.

>>>>

\begin{table}[ht]
    \caption{Sample output from SQL query showing percentage of delivery
    attempts rejected by each restriction}
    \empty{}\label{Sample output from SQL query}
    \begin{tabular}[]{lrr}
        \tabletopline{}%
        Restriction & Number of  & Percentage of    \\
                    & rejections & total rejections \\
        \tablemiddleline{}%
        \input{build/include-sample-sql-output}
        \tablebottomline{}%
    \end{tabular}

    All percentages above are exclusive, i.e.\ the first restriction's
    number of rejections does not include the second restriction's number
    of rejections.

@@

This example client is so poorly written that not only does it present the
sender and recipient addresses improperly, it ignores the error messages
returned by the server and carries on regardless.  Many spam and virus
sending programs have serious deficiencies; unfortunately, other programs,
particularly newer programs, were written by competent programmers or send
mail using well written programs (e.g.\ Postfix or Sendmail on Unix hosts,
Microsoft Outlook on Windows hosts).  Traditionally a mail server would
have done its best to deal with deficient clients, with the intention of
accepting as much mail sent to its users as possible, e.g.\ by ignoring the
absence of a HELO command, or by accepting sender or recipient addresses
that were not enclosed in \texttt{<>}.

### 2009-03-09 doc/logparser-introduction.tex
Corrections based on comments from Carl, plus a few I noticed in the process.

Most mail server administrators will have performed some basic processing
of the log files produced by their mail server at one time or another,
whether it was to debug a problem, explain to a user why their mail is
being rejected, or check if new anti-spam measures are working.  The more
adventurous will have generated statistics to show how successful each of
their anti-spam measures has been in the last week, and possibly even
generated some graphs to clearly illustrate these statistics to management
or users.\footnote{This was the author's first real foray into processing
Postfix log files.}  Very few will have performed in-depth parsing and
analysis of their log files, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing log lines
independently.  One of the barriers to this kind of processing is the
unstructured nature of Postfix log files, where each log line was added on
an ad hoc basis as a requirement was discovered or new functionality was
added.\footnote{A history of all changes made to Postfix is distributed
with the source code, available from
\urlLastChecked{http://www.postfix.org/}{2009/02/23}.}  Further
complication arises because the set of rejection messages is not fixed: new
messages can be added by the administrator with custom checks; every
\acronym{DNSBL}\footnote{This thesis is supplied with a glossary
(\textsection\ref{Glossary}) and a list of acronyms
(\textsection\ref{Acronyms}).} returns a different explanatory message;
policy servers may log different messages depending on the characteristics
of the connection.  There are many ways in which the log lines may differ
between servers, even within the same organisation: servers may be
configured differently, or running different versions of Postfix.

@@

@@

This algorithm requires a database for storing both the rules used when
parsing and the results gleaned from parsing.  The database schema used is
described in \sectionref{database schema}, explaining in detail the tables
used for storing the data gleaned from the log files and the table that
stores the rules.

Chapter~\ref{rules in architecture} discusses the parsing rules in detail,
explaining the purpose and usage of each field in a rule, referring to an
example rule and sample data it matches successfully against.  The pros and
cons of overlapping rules are considered, including techniques for
detecting unintentional overlaps.  Rule efficiency concerns are discussed,
in particular the optimisations used by the algorithm.  The section
concludes with a description of using the tools provided with the parser to
generate new rules (specifically the regex in each rule) from unparsed log
lines.

Chapter~\ref{Postfix Parser Implementation} contains the core of the paper,
describing a naive parsing algorithm and the complications initially
encountered that shaped the full algorithm.  A flow chart and a discussion
of the emergent behaviour exhibited by the algorithm accompanies a
comprehensive explanation of the different stages of the initial algorithm.
The framework that actions and rules fit into is documented, then the
actions taken during execution of the algorithm are described, followed by
the process of adding a new action.  The section concludes with an in-depth
description of the further complications discovered, and their solutions
that complete the parser.

@@

Chapter~\ref{limitations and improvements in implementation} lists the
limitations of the algorithm, then suggests some ways of dealing with them,
with the goal of improving parsing and reproduction of the journey a mail
takes through Postfix.

@@

The bibliography (appendix~\ref{bibliography}) contains references to the
resources used in developing the algorithm, writing the program, and
preparing this thesis.  Also listed are some additional resources expected
to be helpful in understanding \acronym{SMTP}, Postfix, anti-spam
techniques, or the thesis.

>>>>

Most mail server administrators will have performed some basic processing
of the log files produced by their mail server at one time or another,
whether it was to debug a problem, explain to a user why their mail is
being rejected, or check if new anti-spam measures are working.  The more
adventurous will have generated statistics to show how successful each of
their anti-spam measures has been in the last week, and possibly even
generated some graphs to clearly illustrate these statistics to management
or users.\footnote{This was the first real foray the author, a Systems
Administrator for a network of over 2,000 computers and over 1,800 users,
took into processing Postfix log files.}  Very few will have performed
in-depth parsing and analysis of their log files, where the parsing must
correlate the log lines per-connection or per-queueid rather than
processing log lines independently.  One of the barriers to this kind of
processing is the unstructured nature of Postfix log files, where each log
line was added on an ad hoc basis as a requirement was discovered or new
functionality was added.\footnote{A history of all changes made to Postfix
is distributed with the source code, available from
\urlLastChecked{http://www.postfix.org/}{2009/02/23}.}  Further
complication arises because the set of rejection messages is not fixed: new
messages can be added by the administrator with custom checks; every
\acronym{DNSBL}\footnote{This thesis is supplied with a glossary
(\textsection\ref{Glossary}) and a list of acronyms
(\textsection\ref{Acronyms}).} returns a different explanatory message;
policy servers may log different messages depending on the characteristics
of the connection.  There are many ways in which the log lines may differ
between servers, even within the same organisation: servers may be
configured differently, or running different versions of Postfix.

@@

Chapter~\ref{parser architecture} describes the parser architecture
developed for this project, beginning with an overview of the architecture,
then describing the components of the architecture in detail.

@@

Chapter~\ref{Postfix Parser Implementation} discusses the parsing rules in
detail, explaining the purpose and usage of each field in a rule, referring
to an example rule and sample data it matches successfully against.  The
pros and cons of overlapping rules are considered, including techniques for
detecting unintentional overlaps.  Rule efficiency concerns are discussed,
in particular the optimisations used by the algorithm.  The section
concludes with a description of using the tools provided with the parser to
generate new rules (specifically the regex in each rule) from unparsed log
lines.  It contains the core of the paper, describing a naive parsing
algorithm and the complications initially encountered that shaped the full
algorithm.  A flow chart and a discussion of the emergent behaviour
exhibited by the algorithm accompanies a comprehensive explanation of the
different stages of the initial algorithm.  The framework that actions and
rules fit into is documented, then the actions taken during execution of
the algorithm are described, followed by the process of adding a new
action.  The section concludes with an in-depth description of the further
complications discovered, and their solutions that complete the parser.
This algorithm requires a database for storing both the rules used when
parsing and the results gleaned from parsing.  The database schema used is
described in \sectionref{database schema}, explaining in detail the tables
used for storing the data gleaned from the log files and the table that
stores the rules.

@@

@@

The bibliography (appendix~\ref{bibliography}) contains references to the
resources used in developing the algorithm, writing the program, and
preparing this thesis.

### 2009-03-09 doc/logparser-other-parsers-review.tex
Some minor formatting fixes. Documented some of 'make check'. Added check for sentences with only one space separating them.
Removed URLs that are the first Google hit for a search term.
Corrections based on comments from Carl, plus a few I noticed in the process.

There only appears to be one prior published paper about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of \acronym{LMA} is
quite different from \parsername{}: it attempts to present correlated data
from log files in a form suitable for a systems administrator to search
using the myriad of standard Unix text processing utilities already
available.  It produces a \acronym{CSV} file and either a MySQL
(\urlLastChecked{http://www.mysql.com/}{2008/11/23}) or Berkeley DB
(\urlLastChecked{http://www.oracle.com/database/berkeley-db/index.html}{2008/11/23})
database.  The decision to support both \acronym{CSV} and Berkeley DB
appears to have been a serious limitation: both formats have limitations
that will be explained later.  Hardly any documentation is provided with
\acronym{LMA}, though some documentation is available
in~\cite{log-mail-analyser}.  Studying the source code is informative,
though this author had difficulty as the authors of \acronym{LMA} wrote in
Italian.

@@

Whether a MySQL database or Berkeley DB table is chosen in addition to the
\acronym{CSV} output, \acronym{LMA} stores the following data: time and
date, client hostname and \acronym{IP} address, server \acronym{IP}
address, sender and recipient addresses, \acronym{SMTP} code, and size (for
accepted mails only).  It is unclear what time and date is stored: start
time, end time, or delivery time?  Unlike \parsername{} it does not store
the server hostname, HELO hostname, queueid, start and end times,
timestamps for each log line, enhanced status code, delivery delays, or
message id (for accepted mails only).  Handling of multiple recipients,
\acronym{SMTP} codes, or remote servers\footnote{A single mail may be sent
to multiple remote servers if it was addressed to recipients in different
domains, or Postfix needs to try multiple servers for one or more
recipients.} is not explained; experimental observation shows that multiple
records are added when a mail had multiple recipients (sadly the records
are not associated or linked in any way), and presumably the same approach
is taken when there were multiple destination servers.

@@

\acronym{LMA} does not deal with any of the complications discussed in
\sectionref{complications}, except for correlating log lines by queueid;
not correlating log lines by pid means it cannot correlate most rejections.
It does not differentiate between different types of rejections, so it is
not suitable for the purposes of this project; the data about which
restriction caused the rejection is discarded, whereas the main goal of
this project is to retain that data to aid optimisation and evaluation of
anti-spam techniques.  \acronym{LMA} fails to parse Postfix log files
generated on Solaris hosts because the fields automatically prepended to
each log line differ from those added on Linux hosts; log files from
Solaris hosts (and possibly other operating systems) thus require
preprocessing before parsing by \acronym{LMA}.  The \numberOFlogFILES{}
preprocessed test log files were parsed without complaint by \acronym{LMA},
although it produced 32 entries in the output \acronym{CSV} file for every
rejection in the input log file; it also missed some 40\% of delivered
mail.  Once these glaring deficiencies were discovered the author did not
waste any more time checking the results.

\acronym{LMA} does provide some simple reports: IP-STORY, FROM-STORY,
DAILY-EMAIL and DAILY-REJECT\@.  These reports search the Berkeley DB files
for matching records: the first three extract \acronym{CSV} lines for the
specified client \acronym{IP} address, sender address, or date
respectively.  DAILY-REJECT initially failed with an error message from the
Perl interpreter;\footnote{The error messages were:
\newline{}\texttt{Undefined subroutine \&main::LIST called at queryDB.pl
line 372.}\newline{}\texttt{Undefined subroutine \&main::EXTRACT\_FROM\_DB
called at queryDB.pl line 379.}} after making corrections to the code it
worked, extracting the \acronym{CSV} lines for the specified day where the
\acronym{SMTP} code signifies a rejection.  All of these reports are
trivially simple to produce from the \acronym{CSV} file using the standard
Unix tool \texttt{awk}\glsadd{awk}; the most complicated, DAILY-REJECT, is
merely:

@@

    \item DAILY-REJECT considers all \acronym{SMTP} codes except
        \texttt{250} to be rejections; this includes invalid \acronym{SMTP}
        codes such as \texttt{0} and \texttt{deferred}, so the awk command
        does too.  These invalid \acronym{SMTP} codes are most likely
        present because of incorrect parsing by \acronym{LMA}.

@@

Pflogsumm produces a report designed for troubleshooting rather than
in-depth analysis.  It does not support saving any data, and it does not
extract any data that it does not require to produce its report, e.g.\ it
does not extract the HELO hostname, queueid, start and end times,
timestamps for each log line, or message id.  Both the parsing and
reporting are difficult to extend because it is a specialised tool, unlike
the easily extensible design of \parsername{}.  It does not correlate log
lines by queueid or \acronym{pid}, and does not need to deal with the
complications encountered during this project (\sectionref{complications}).
Pflogsumm produces a useful report, and successfully parsed the
\numberOFlogFILES{} log files it was tested with.\footnote{The results it
reported were not verified in detail, but it did not report any errors, and
has an excellent reputation amongst Postfix users.}  Pflogsumm has many
options to include or exclude certain sections of the report, all clearly
documented; by default it includes the following:

@@

\parserblurb{%
    Sawmill is a Postfix log analyzer (it also support 818 other log
    formats). It can process log files in Postfix format, and generate
    dynamic statistics from them, analyzing and reporting events. Sawmill
    can parse Postfix logs, import them into a SQL database (or its own
    built-in database), aggregate them, and generate dynamically filtered
    reports, all through a web interface. Sawmill can perform Postfix
    analysis on any platform, including Window, Linux, FreeBSD, OpenBSD,
    Mac OS, Solaris, other UNIX, and more.
}
{http://www.thesawmill.co.uk/formats/postfix.html}
{2008/11/23}

@@

The source code is available in an obfuscated form, and the product is
quite expensive, requiring a \euros{100} + VAT licence per report
(discounts are available when buying multiple licences); in contrast
\parsername{} is free to use and the code is freely available.  Sawmill is
supplied with thorough and well written documentation; everything the
author searched for was documented, except the MySQL database schema and
some details of the data extracted by the parser.  A commercial version of
MySQL is required because of MySQL licensing restrictions, but Sawmill's
documentation explains why and includes instructions on how to compile
Sawmill so that it can use a non-commercial version of MySQL (this was not
attempted during the review process).

Sawmill's web interface supports searching on any combination of the fields
it extracts, and all searches produced accurate results.  The interface for
searching is neither as simple to use nor as informative as the interface
provided by Splunk.  The administrative interface is much easier to use
than Splunk's: it took only five minutes to start parsing a whole directory
of log files.

@@

In summary Sawmill suffers from being a general purpose product; it is
probably much more useful when parsing log files where each log line is
self-contained (e.g.\ web server log files), rather than log files
containing interlinked log lines.  It is not suitable as a base for this
parser, as the source code made available is obfuscated and not intended
for modification; in addition the architecture would probably need to be
overhauled or replaced to deal with correlating log lines.

@@

@@

The web interface is quite attractive and simple to use when searching, but
as an administrator it seems unnecessarily difficult to perform simple
tasks.  When testing Splunk it took roughly 30 minutes to figure out how to
add a single log file to be indexed so that it could be searched, with the
downside that the log file was copied into a spool directory before
indexing, doubling the disk space usage.  The next test was to index all
the log files in a particular directory, but after three hours, numerous
futile attempts, and reading all the available documentation, the author
admitted defeat.  Using the \acronym{CLI} rather than the web interface was
partially successful: the command \newline{} \tab{} \texttt{splunk find
logs }\textit{log-directory\/}\newline{} added 40 of the
\numberOFlogFILES{} log files to the queue for indexing.  Further attempts
enqueued the same 40 log files, without explaining why the others were
excluded.\footnote{The log files appear to have been indexed once only;
presumably Splunk keeps track of the log files it has indexed and discards
requests to index log files for a second time.  This may or may not be a
useful feature for \parsername{}.} There did not appear to be an option to
ensure the log files would be processed in the order they were created,
though this may be neither necessary nor beneficial with Splunk.
Subsequently the author was successful in adding a single file at a time
using the \acronym{CLI}:\newline{} \tab{} \texttt{splunk add tail
}\textit{filename\/}\newline{} A simple loop was then enough to add all the
desired log files.  Splunk will periodically check all indexed log files
for updates unless they are manually removed from its list; this may or may
not be useful behaviour.  Splunk did not appear to have any difficulty in
indexing the log files, once they had been successfully added to its queue.
\parsername{} parses the logs it is instructed to parse, in the order
given; periodic parsing of logs is a task an administrator can easily
achieve with \texttt{cron(8)} and \texttt{logrotate(8)}.

Copious documentation is made available on
\urlLastChecked{http://www.splunk.com/}{2008/11/23}, but poor organisation
and sheer abundance makes it extremely hard to find useful information.
Searches confusingly tended to return results from old documentation rather
than new.  In general the documentation appears to have been written by
someone intimately acquainted with the software, who has difficulty
understanding how a newcomer would approach tasks or the questions they
would ask.

Splunk supports reading compressed log files without any configuration by
the user.  The free version of Splunk limits the volume of data indexed per
day to 500MB, though a trial Enterprise licence is available that allows
indexing of up to 5GB of data per day.  In 2007 the cheapest licenced
version cost \$5000 plus \$1000 support, and limited the volume of data
indexed per day to 500MB\@.  Prices were removed from the Splunk website
during 2008; now Splunk's sales team must be asked for a quote.  Typical
log file sizes for a small scale mail server are given in
\sectionref{parser efficiency}.

@@

AWStats can produce simple graphs from many different services' log files,
but supporting numerous log files formats without special purpose code
limits its functionality.  The data it can extract from an \acronym{MTA}
log file is limited in comparison to \parsername{}: time2, email, email\_r,
host, host\_r, method, url, code, and bytesd.  No explanation for any of
those fields is provided in the documentation at
\urlLastChecked{http://awstats.sourceforge.net/docs/awstats_faq.html\#MAIL}{2008/11/23}
(in contrast \parsername{} fully documents everything it extracts), so the
author could not understand what data is extracted, or determine what data
is missing in comparison to \parsername{}.  AWStats coerces Postfix log
files into Apache\footnote{The Apache web server is the most popular HTTP
server in use over the past 10 years; more information is available at
\urlLastChecked{http://httpd.apache.org/}{2008/11/23}.} format log files,
for analysis by AWStats' HTTP log file parser.  The converting parser only
deals with a small portion of the log lines generated by Postfix, silently
skipping those it cannot deal with, and does not distinguish between
different types of rejection; extending it to parse all log lines would be
at least as much work as writing a new parser.  It does correlate log lines
by queueid (not by pid), but it does not deal with any of the other
complications described in \sectionref{complications}.  AWStats supports
saving data, but the format of the data store is not documented.  It also
supports reading compressed log files, but that functionality was not
tested.

When tested with the \numberOFlogFILES{} test log files AWStats' reported
that it parsed 9,240,075 (88.70\%) of 10,416,129 log lines, skipping
1,176,050 (11.29\%) corrupt log lines; the \numberOFlogFILES{} log files
contain \numberOFlogLINES{} log lines, so AWStats parsed only 15.21\% of
the log lines, declared 1.93\% were corrupt, and ignored the remaining
82.85\%.

@@

Anteater does not have any English documentation so it is impossible for
this author to accurately comment on the analysis it performs.  It did not
run successfully when tested, and its parsing would certainly be out of
date because Postfix has evolved considerably since this tool was last
updated (2003/11/06).  As it neither ran successfully nor has documentation
the author can read, a detailed review cannot be provided.

The Debian Linux distribution
(\urlLastChecked{http://www.debian.org/}{2008/11/23}) provides a manual
page with the copy of anteater it distributes, so the author was at least
able to run anteater with the correct arguments; sadly anteater produced
zero for every statistic, presumably because it was unsuccessful in parsing
the log lines.

@@

YAALA uses a plugin-based system to analyse log files and produce reports
in HTML format, with all the parsing and report generation handled by
plugins.  Using YAALA as a base for this project would be only slightly
less work than starting from scratch, as both the input and output modules
would need to be written specially; it might be more work to implement a
parser within the constraints of YAALA\@.  YAALA supports storing
previously gathered data using Perl's Storable module, so other Perl
programs can use Storable to load, examine, and optionally modify the data;
\parsername{} uses a well documented database that is accessible from most
common programming languages.  This information was gleaned from the source
code, as the documentation is sadly lacking.

YAALA provides a Postfix parser that extracts the following types of fields
from specific log lines:

@@

In summary YAALA provides a Postfix parser that tries to parse the most
common Postfix log lines only, provides reasonably flexible report
generation from the limited data extracted, but has no facilities to
extract data for use in other tools.

@@

Lire supports multiple output formats for generated reports (text, HTML,
PDF, and Excel 95) but the reports do not appear to be customisable;
\parsername{} does not produce any reports.  Lire's report is not as
detailed as Pflogsumm's, and it is considerable harder to configure.  Lire
supports saving extracted data for later report generation, but the format
of this data store is undocumented; given the source code it should be
possible, with enough time and effort, to understand the format.
\parsername{} uses an \acronym{SQL} database to make accessing the
extracted data as easy as possible.  In general Lire has poor
documentation.

>>>>

There only appears to be one prior published paper about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of \acronym{LMA} is
quite different from \parsername{}: it attempts to present correlated data
from log files in a form suitable for a systems administrator to search
using the myriad of standard Unix text processing utilities already
available.  It produces a \acronym{CSV} file and either a MySQL or Berkeley
DB database.  The decision to support both \acronym{CSV} and Berkeley DB
appears to have been a serious limitation: both formats have limitations
that will be explained later.  Hardly any documentation is provided with
\acronym{LMA}, though some documentation is available
in~\cite{log-mail-analyser}.  Studying the source code is informative,
though this author had difficulty as the authors of \acronym{LMA} wrote in
Italian.

@@

Whether a MySQL database or Berkeley DB table is chosen in addition to the
\acronym{CSV} output, \acronym{LMA} stores the following data: time and
date of the log line, client hostname and \acronym{IP} address, server
\acronym{IP} address, sender and recipient addresses, \acronym{SMTP} code,
and size (for accepted mails only).  Unlike \parsername{} it does not store
the server hostname, HELO hostname, queueid, start and end times,
timestamps for each log line, enhanced status code, delivery delays, or
message id (for accepted mails only).  Handling of multiple recipients,
\acronym{SMTP} codes, or remote servers\footnote{A single mail may be sent
to multiple remote servers if it was addressed to recipients in different
domains, or Postfix needs to try multiple servers for one or more
recipients.} is not explained; experimental observation shows that multiple
records are added when a mail had multiple recipients (sadly the records
are not associated or linked in any way), and presumably the same approach
is taken when there were multiple destination servers.

@@

\acronym{LMA} does not deal with any of the complications discussed in
\sectionref{complications}, except for correlating log lines by queueid;
not correlating log lines by pid means it cannot correlate most rejections.
It does not differentiate between different types of rejections, so it is
not suitable for the purposes of this project; the data about which
restriction caused the rejection is discarded, whereas the main goal of
this project is to retain that data to aid optimisation and evaluation of
anti-spam techniques.  \acronym{LMA} fails to parse Postfix log files
generated on Solaris hosts because the fields automatically prepended to
each log line differ from those added on Linux hosts; log files from
Solaris hosts (and possibly other operating systems) thus require
preprocessing before parsing by \acronym{LMA}.  The \numberOFlogFILES{}
preprocessed test log files were parsed without complaint by \acronym{LMA},
although it produced 32 entries in the output \acronym{CSV} file for every
rejection in the input log file; it also missed some 40\% of delivered
mail.  Once these deficiencies were discovered the author did not spend any
more time checking the results.

\acronym{LMA} does provide some simple reports: IP-STORY, FROM-STORY,
DAILY-EMAIL and DAILY-REJECT\@.  These reports search the Berkeley DB files
for matching records: the first three extract \acronym{CSV} lines for the
specified client \acronym{IP} address, sender address, or date
respectively.  DAILY-REJECT initially failed with an error message from the
Perl interpreter;\footnote{The error messages were:
\newline{}\texttt{Undefined subroutine \&main::LIST called at queryDB.pl
line 372.}\newline{}\texttt{Undefined subroutine \&main::EXTRACT\_FROM\_DB
called at queryDB.pl line 379.}} after correcting the errors in the code it
worked, extracting the \acronym{CSV} lines for the specified day where the
\acronym{SMTP} code signifies a rejection.  All of these reports are
trivially simple to produce from the \acronym{CSV} file using the standard
Unix tool \texttt{awk}\glsadd{awk}; the most complicated, DAILY-REJECT, is
merely:

@@

    \item DAILY-REJECT considers all \acronym{SMTP} codes except ``250'' to
        be rejections; this includes invalid \acronym{SMTP} codes such as
        \texttt{0} and \texttt{deferred}, so the awk command does too.
        These invalid \acronym{SMTP} codes are most likely present because
        of incorrect parsing by \acronym{LMA}.

@@

Pflogsumm produces a report designed for troubleshooting rather than
in-depth analysis.  It does not support saving any data, and it does not
extract any data that it does not require to produce its report; e.g.\ it
does not extract the HELO hostname, queueid, start and end times,
timestamps for each log line, or message id.  Both the parsing and
reporting are difficult to extend because it is a specialised tool, unlike
the easily extensible design of \parsername{}.  It does not correlate log
lines by queueid or \acronym{pid}, and does not need to deal with the
complications encountered during this project (\sectionref{complications}).
Pflogsumm produces a useful report, and successfully parsed the
\numberOFlogFILES{} log files it was tested with.\footnote{The results it
reported were not verified in detail, but it did not report any errors, and
has an excellent reputation amongst Postfix users.}  Pflogsumm has many
options to include or exclude certain sections of the report, all clearly
documented; by default it includes the following:

@@

\parserblurb{%
    Sawmill is a Postfix log analyzer (it also support 818 other log
    formats).  It can process log files in Postfix format, and generate
    dynamic statistics from them, analyzing and reporting events.  Sawmill
    can parse Postfix logs, import them into a SQL database (or its own
    built-in database), aggregate them, and generate dynamically filtered
    reports, all through a web interface.  Sawmill can perform Postfix
    analysis on any platform, including Window~[sic], Linux, FreeBSD,
    OpenBSD, Mac OS, Solaris, other UNIX, and more.
}
{http://www.thesawmill.co.uk/formats/postfix.html}
{2008/11/23}

@@

The source code is only available in an encrypted form, to support people
who wish to use Sawmill on operating systems or machine architectures the
company do not provide executables for.  Sawmill is quite expensive,
requiring a \euros{100} + VAT licence per report, with discounts available
when buying multiple licences (correct as of 2008/11/23); in contrast,
\parsername{} is free to use and the code is freely available.  Sawmill is
supplied with thorough and well written documentation; everything the
author searched for was documented, except the MySQL database schema and
some details of the data extracted by the parser.  A commercial version of
MySQL is required because of MySQL licensing restrictions, but Sawmill's
documentation explains why and includes instructions on how to compile
Sawmill so that it can use a non-commercial version of MySQL (this was not
attempted during the review process).

Sawmill's web interface supports searching on any combination of the fields
it extracts, and all searches produced accurate results.  The interface for
searching is neither as simple to use nor as informative as the interface
provided by Splunk (see \sectionref{Splunk review}).  The administrative
interface is much easier to use than Splunk's: it took only five minutes to
start parsing a whole directory of log files.

@@

In summary, Sawmill suffers from supporting so many types of log files: it
parses many types of log files, but none of them very well; it is probably
much more useful when parsing log files where each log line is
self-contained (e.g.\ web server log files), rather than log files
containing interlinked log lines.  It is not suitable as a base for this
parser, as the source code made available is obfuscated and not intended
for modification; in addition the architecture would probably need to be
overhauled or replaced to deal with correlating log lines.

@@

\label{Splunk review}

@@

The web interface is quite attractive and simple to use when searching, but
as an administrator it seems unnecessarily difficult to perform simple
tasks.  When testing Splunk it took roughly 30 minutes to figure out how to
add a single log file to be indexed so that it could be searched, with the
downside that the log file was copied into a spool directory before
indexing, doubling the disk space usage.  The next test was to index all
the log files in a particular directory, but after three hours, numerous
futile attempts, and reading all the available documentation, the author
was still unable to index all the log files in a directory using the web
interface.  Using the \acronym{CLI} rather than the web interface was
partially successful: the command ``\texttt{splunk find logs
}\textit{log-directory\/}'' added 40 of the \numberOFlogFILES{} log files
to the queue for indexing.  Further attempts enqueued the same 40 log
files, without explaining why the others were excluded.\footnote{The log
files appear to have been indexed once only; presumably Splunk keeps track
of the log files it has indexed and discards requests to index log files
for a second time.  This may or may not be a useful feature for
\parsername{}.} There did not appear to be an option to ensure the log
files would be processed in the order they were created, though this may be
neither necessary nor beneficial with Splunk.  Subsequently the author was
successful in adding a single file at a time using the \acronym{CLI} and
the command ``\texttt{splunk add tail }\textit{filename\/}''.  A simple
loop using that command was enough to add all the desired log files.
Splunk will periodically check all indexed log files for updates unless
they are manually removed from its list; this may or may not be useful
behaviour.  Splunk did not appear to have any difficulty in indexing the
log files, once they had been successfully added to its queue.
\parsername{} parses the logs it is instructed to parse, in the order
given; periodic parsing of logs is a task an administrator can easily
achieve with \texttt{cron(8)} and \texttt{logrotate(8)}.

Copious documentation is made available on
\urlLastChecked{http://www.splunk.com/}{2008/11/23}, but the organisation
and abundance of material does not support easy identification of useful
information.  Searches confusingly tended to return results from old
documentation rather than new.  In general the documentation appears to
have been written by someone intimately acquainted with the software, who
has difficulty understanding how a newcomer would approach tasks or the
questions they would ask.

Splunk supports reading compressed log files without any configuration by
the user.  The free version of Splunk limits the volume of data indexed per
day to 500MB, though a trial Enterprise licence is available that allows
indexing of up to 5GB of data per day.  In 2007, the cheapest licenced
version cost \$5000 plus \$1000 support, and limited the volume of data
indexed per day to 500MB\@.  Prices were removed from the Splunk website
during 2008; now Splunk's sales team must be asked for a quote.  Typical
log file sizes for a small scale mail server are given in
\sectionref{parser efficiency}.

@@

AWStats can produce simple graphs from many different services' log files,
but supporting numerous log files formats without special purpose code
limits its functionality.  The data it can extract from an \acronym{MTA}
log file is limited in comparison to \parsername{}: time2, email, email\_r,
host, host\_r, method, url, code, and bytesd.  No explanation for any of
those fields is provided in the documentation at
\urlLastChecked{http://awstats.sourceforge.net/docs/awstats_faq.html\#MAIL}{2008/11/23},
so the author could neither understand what data is extracted, nor
determine what data is missing in comparison to \parsername{} (which fully
documents everything it extracts).  AWStats coerces Postfix log files into
the log file format used by the Apache web server, for analysis by AWStats'
HTTP log file parser.  The converting parser only deals with a small
portion of the log lines generated by Postfix, silently skipping those it
cannot deal with, and does not distinguish between different types of
rejection; extending it to parse all log lines would be at least as much
work as writing a new parser.  It does correlate log lines by queueid (not
by pid), but it does not deal with any of the other complications described
in \sectionref{complications}.  AWStats supports saving data, but the
format of the data store is not documented.  It also supports reading
compressed log files, but that functionality was not tested.

When tested with the \numberOFlogFILES{} test log files AWStats' reported
that it parsed 9,240,075 (88.70\%) of 10,416,129 log lines, skipping
1,176,050 (11.29\%) corrupt log lines; the \numberOFlogFILES{} log files
contain \numberOFlogLINES{} log lines, so AWStats parsed only 15.21\% of
the log lines, declared 1.93\% were corrupt, and ignored the remaining
82.85\%.  The author did not verify the correctness of the parsing of those
log lines AWStats did parse.

@@

Anteater does not have any English documentation except for the quote above
so it is impossible for this author to accurately comment on the analysis
it performs.  It did not run successfully when tested, and its parsing
would certainly be out of date because Postfix has evolved considerably
since this tool was last updated (2003/11/06).  As it neither ran
successfully nor has documentation the author can read, a detailed review
cannot be provided.

The Debian Linux distribution provides a translated manual page with the
copy of anteater it distributes, so the author was at least able to run
anteater with the correct arguments; sadly anteater produced zero for every
statistic, presumably because it was unsuccessful in parsing the log lines.

@@

YAALA uses a plugin-based system to analyse log files and produce reports
in HTML format, with all the parsing and report generation handled by
plugins.  Using YAALA as a base for this project would have been as much
work as starting from scratch, as both the input and output modules would
need to be written specially; it might be more work to implement a parser
within the constraints of YAALA\@.  YAALA supports storing previously
gathered data using Perl's Storable module, so other Perl programs can use
Storable to load, examine, and optionally modify the data; \parsername{}
uses a well documented database that is accessible from most common
programming languages.  Information about how YAALA stores data was gleaned
from the source code, as the documentation is sadly lacking.

YAALA provides a Postfix parser that extracts the following of fields from
specific log lines:

@@

In summary, YAALA provides a Postfix parser that tries to parse the most
common Postfix log lines only, provides reasonably flexible report
generation from the limited data extracted, but has no facilities to
extract data for use in other tools.

@@

Lire supports multiple output formats for generated reports (text, HTML,
PDF, and Excel 95) but the reports do not appear to be customisable;
\parsername{} does not produce any reports.  Lire's report is not as
detailed as Pflogsumm's, and it is considerable harder to configure.  Lire
supports saving extracted data for later report generation, but the format
of this data store is undocumented; given the source code it should be
possible, with enough time and effort, to understand the format.
\parsername{} uses an \acronym{SQL} database to make accessing the
extracted data as easy as possible.  In general, Lire has poor
documentation.
